Reproducibility of SNP-calling in multiple sequencing runs from single tumors

### 2015-06-02 PAPER/paper.tex
Matt's edits are included in the source file.
I think the paper is MUCH improved over the previous version.  Well done!

My main comment is that I would not call somatic mutations as SNPs

We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling. We used the same computational pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates. We further tested whether additional filtering increased or decreased the size of the overlap.  We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966). We further found that using filtering after SNP-calling removed the overlap completely. We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.

@@

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large scale genomic data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

However, these and other efforts have produced very few results for patients in terms of improved treatments. It may be that we simply haven't yet analyzed the data in the most informative manner. Another factor may be that the data are noisy. It is well known that sequencing and SNP-calling pipelines are not error-free. For example, different SNP-calling pipelines can return different results on the same data \citep{SNPcall}. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low. Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNP-calls. Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return comparable results on two sequencing runs from the same tissue. And further, we asked whether added filtering after SNP-calling increases or decreases the degree of similarity among replicates.

@@

% Experiment: size of overlap (figures 3 and 4)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}. Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or completely absent. In general, we would like to know how often these errors occur. One way of investigating this question would be to use PCR to verify each individual mutation. However, this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates. Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not. (Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.

TCGA's GBM data set includes 2 technical replicates each of 55 tumors. In this case, the technical replicates are not identical. One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate. Despite the difference in sequencing protocol, we would expect any fully penetrant mutation to appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one. Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate. We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNPs per sample, than the corresponding WGS samples.

@@

% Experiment: which filters do what (figures 6 and 7) 
% Experiment: those two filters are different (figures 8 and 9)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}. Software platforms to perform these tasks are plentiful, and each one typically employs multiple methods to identify true positive mutations. The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and putative mutations with higher quality scores are considered to be more likely true somatic mutations than those with low quality scores. In addition to considering these quality scores, there are additional filtering steps that one can use to distinguish true somatic mutations from errors of all sorts. Table~\ref{tab:filters} lists eight distinct SNP filters we evaluated. The first three are based on the quality scores generated by GATK and SomaticSniper.  We simply remove from the dataset anything that fails to meet conventional quality-score thresholds. The other five are additional filters that we developed for this project. Each of these five filters represents an aspect of the data or putative mutation that is generally thought to indicate that a given SNP call is a false positive.

@@

GBM is an evolutionary disease that develops when mutations arise in glial cell lines and the mutated cells and their lineages subsequently co-opt the surrounding tissue and systems to the detriment of the organism as a whole. Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but may be improved by a more complete understanding of the somatic mutations present in GBM. Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers. These data are enormous potential value to the research community, but their accuracy and reproducibility are unknown. Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.

We found that on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates. The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations. We found further that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates. Moreover, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations. These results suggest the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.

>>>>

We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling. We used the same mutation-calling pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates. We further tested whether additional filtering increased or decreased the size of the overlap.  We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966). We further found that using filtering after SNP-calling removed the overlap completely. We concluded that there is variation in the frequency of mutations in GBMs, and that while some filtering approaches preferentially removed putative mutations found in only one replicate, others removed a large fraction of putative mutations found in both.

@@

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

% Matt: I would use SNV, rather than SNP, since SNP is more commonly used to refer to germline variants.  Or just use "somatic mutation calling pipeline"
However, these and other efforts have produced very few results for patients in terms of improved treatments. It may be that we simply haven't yet analyzed the data in the most informative manner. Another factor may be that the data are noisy. It is well known that sequencing and SNP-calling pipelines are not error-free. For example, different SNP-calling pipelines can return different results on the same data \citep{SNPcall}. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low. Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNP-calls. Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return comparable results on two sequencing runs from the same tissue. And further, we asked whether added filtering after SNP-calling increases or decreases the degree of similarity among replicates.

@@

% Experiment: size of overlap (figures 3 and 4)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
%
% Matt:
%  a.  LOH can be real or an amplification artifact
%  b.  isn't penetrance more about phenotype?  i would use polymorphic (for partially penetrant) and fixed (for fully penetrant)
%  c.  i think the main reason for lack of multi-sample sequencing is that TCGA and other have used pathology specimens (i.e. there is only one per patient) out of convenience

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs and by mis-aligned bases during sequence analysis \citep{seqerror}. Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or completely absent. In general, we would like to know how often these errors occur. One way of investigating this question would be to use an orthogonal technique such as PCR to verify each individual mutation. However, this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates. Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not. (Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiments, but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.

% Matt:
%  a.  the first citation looks weird in the PDF.  it says (Network, 2008), which I think should be (TCGA Network, 2008)

TCGA's GBM data set includes 2 technical replicates for each of 55 tumors. In this case, the technical replicates are not identical. One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate. Despite the difference in sequencing protocol, we would expect any fully penetrant mutation to appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one. Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate. We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNPs per sample, than the corresponding WGS samples.

@@

% Experiment: which filters do what (figures 6 and 7) 
% Experiment: those two filters are different (figures 8 and 9)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
As an additional computational validation step for somatic mutations, it is common practice to employ computational algorithms that attempt to distinguish somatic mutations from germ line mutations and sequencing errors \citep{SomaticSniper, mut_calling}. Software platforms to perform these tasks are plentiful, and each one typically employs multiple methods to identify true positive mutations. The two platforms used in this research, SomaticSniper \citep{SomaticSniper} and GATK \citep{GATK}, calculate one or more quality scores based on features of the dataset and the individual reads, and putative mutations with higher quality scores are considered to be more likely true somatic mutations than those with low quality scores. In addition to considering these quality scores, there are additional filtering steps that one can use to distinguish true somatic mutations from errors of all sorts. Table~\ref{tab:filters} lists eight distinct SNP filters we evaluated. The first three are based on the quality scores generated by GATK and SomaticSniper.  We simply remove from the dataset anything that fails to meet conventional quality-score thresholds. The other five are additional filters that we developed for this project. Each of these five filters represents an aspect of the data or the putative mutation that is generally thought to indicate that a given SNP call is a false positive.

@@

GBM is an evolutionary disease that develops when mutations arise in glial cell lines and the mutated cells and their lineages subsequently co-opt the surrounding tissue and systems to the detriment of the organism as a whole. Treatment for GBM is difficult and has poor outcomes \citep{GBM-stats}, but may be improved by a more complete understanding of the somatic mutations present in GBM. Large-scale sequencing projects, like TCGA, make cancer sequencing data available to many researchers. These data are of enormous potential value to the research community, but their accuracy and reproducibility are unknown. Here, we have made a first step towards evaluating the reliability of the TCGA data, by comparing 55 technical replicates in the TCGA GBM dataset.

We found that, on average, about half of the putative mutations in the raw data for the WGS replicate (no amplification before library preparation) and about a third of those in the WGA replicate (with amplification before library preparation) were present in both replicates. The number of mutations present in both replicates was anywhere between 20 and 5,000 putative mutations. We found further that the high number of putative somatic mutations in some, but not all, of the patient samples was repeatable across technical replicates. Moreover, samples with a higher frequency of putative mutations had equally similar technical replicates to those samples with a lower frequency of putative mutations. These results suggest the possibility that a higher mutation frequency could be a feature of a subset of GBM tumors and not a data artifact.
